The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2012-05-29 smurf_paper.tex
work on intro, new section on convergence tests

% journals
\newcommand{\apj}{\rm ApJ}
\newcommand{\apjl}{\rm ApJL}
\newcommand{\apjs}{\rm ApJS}
\newcommand{\aaps}{\rm A$\&$AS}
\newcommand{\aap}{\rm A$\&$A}
\newcommand{\aapr}{\rm A$\&$AR}
\newcommand{\mnras}{\rm MNRAS}
\newcommand{\aj}{\rm Astron. J.}
\newcommand{\araa}{\rm ARAA}
\newcommand{\nat}{\rm Nature}
\newcommand{\pasj}{\rm PASJ}
\newcommand{\pasp}{\rm Publ. Astron. Soc. Pac.}
\newcommand{\prd}{\rm PhRvD}
\newcommand{\ASP}{\rm ASP Conference Series}
\newcommand{\CASP}{\rm Comm. Astrophys. Space Phys.}
\newcommand{\astroph}{\rm astro-ph/}

@@

\author[Edward~L.~Chapin~et~al.]{
  \parbox[t]{\textwidth}{
    Edward~L.~Chapin$^{1,2}$\thanks{E-mail:~echapin@phas.ubc.ca},
    David~S.~Berry$^{2}$,
    Andrew~G.~Gibb$^{1}$,
    Tim~Jenness$^{2}$,
    Douglas~Scott$^{1}$,
    Frossie~Economou$^2$\thanks{Present address: National Optical
      Astronomy Observatory, 950 N.\ Cherry Avenue, Tucson, AZ 85719, USA}
  }
  \\
  \\
  $^{1}$Dept. of Physics \& Astronomy, University of British Columbia,
  6224 Agricultural Road, Vancouver, B.C. V6T 1Z1, Canada\\
  $^{2}$JointAstronomy Centre, 660 N. A`oh\={o}k\={u} Place, University
  Park, Hilo, Hawaii 96720, USA}

@@

\textbf{Guideline for figures:}

\begin{itemize}
\item minimize excess whitespace around the figure to optimize usage of space
\item ensure a decent line thickness
\item use a Times font for labels with a thickness that resembles the
  text in the body. The font size should be similar to the main body
  text or larger
\item avoid using colour unless it is necessary for emphasis
\end{itemize}

\textbf{start intro here...}

@@

Over the last twenty years, observations in the submillimetre band
(defined here to be 200--1000\,\micron) have helped revolutionized
several important areas of astrophysics including: the
characterization of early stages of star-formation by identifying the
dense, cold regions in molecular clouds where stars may eventually
form; discovering through blind surveys a class of massive dusty
star-forming galaxies in the early ($z>2$) Universe, now referred to
as submillimetre galaxies, or SMGs; and identifying debris disks
around nearby stars, the early stages of planet formation.  With
$\sim10000$ total detectors, SCUBA-2 is presently the largest
submillimetre camera in the world, and the fastest wide-area
ground-based submillimetre imager.

Most submillimetre imaging cameras to date have used bolometers to
maximize sensitivities: each pixel consists of an efficient thermal
absorber to collect photons that have passed through a broad-band
filter, and the signal is read-out using a thermometer. The
high-frequency broad-band noise in modern systems, such as SCUBA-2, is
limited by photon noise from the instrument and ambient
backgrounds. The low-frequency noise, however, is dominated by sources
such as the low-frequency drift in the readout electronics, and drifts
in the mean background levels (such as the atmosphere for ground-based
instruments). The process of making a map therefore amounts to
separating the component of the bolometer signals that is produced by
astronomical sources, from the non-white (and therefore correlated)
contaminating noise components. Fortunately each has a particular
profile: astronomical sources are typically fixed in a sky reference
frame (to a good approximation), whereas other noise sources are often
correlated in time, and usually from bolometer-to-bolometer. ...also
mention scan strategies...

There are numerous ways to attack this problem, most of them evolving
out of techniques that were developed to produce maps of the
large-scale anisitropies in the Cosmic Microwave Background (CMB).
Styles of map-making: least squares \citep{janssen1992} maximum
likelihood \citep[e.g.,][]{patanchon2008}; iterative -- see
\citet{johnstone2000} implementation of \citet{wright1996} to remove
chop, \citet{kovacs2008} for SHARC-2 etc.; de-correlation using PCA
analysis \citep[e.g.][]{laurent2005,scott2008,aguirre2010}.

@@

We then express the noise, $\mathbf{n}_i(t)$, as the sum of several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) + \mathbf{n}^\mathrm{f}_i(t) +
  \mathbf{n}^\mathrm{c}(t),
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white (thermal)
noise, $\mathbf{n}^\mathrm{f}_i(t)$ is low-frequency noise that is
\emph{not} correlated from bolometer-to-bolometer, and
$\mathbf{n}^\mathrm{c}(t)$ is a (predominantly low-frequency)
correlated or \emph{common-mode} component. The primary goal of
map-making is to model and remove $\mathbf{n}^\mathrm{f}_i$ and
$\mathbf{n}^\mathrm{c}$ from the time-series, while re-gridding the
remaining data we can then hope to approach the theoretical noise
limit in the map. This procedure is summarized in Fig.~\ref{fig:dimm}.

@@

In addition, a variance map $\mathbf{v}(x_j,y_j)$ is estimated. The
default behaviour is to estimate this weighted error on the mean given
the scatter in the weighted samples. This is accomplished by dividing
the \emph{biased weighted sample variance} by the number of samples
that went into the average (akin to the formula for standard error on
the mean, but accounting for weights),
%
\begin{equation}
\mathbf{v}(x_j,y_j) = \frac{\sum_j \mathbf{w}_j
                            \sum_j \mathbf{w}_j \mathbf{b}_j -
                            \left( \sum_j \mathbf{w}_j \mathbf{b}_j \right)^2 }
                           { N \left( \sum_j \mathbf{w}_j \right)^2 },
\end{equation}
%
where $N_j$ is the total number of bolometer samples that land in the
pixel. As written, this algorithm is numerically unstable if the two
terms in the numerator are large, and of nearly the same value:
floating point precision errors can cause the difference to be
significantly incorrect. In practice we use the more numerically
stable ``weighted incremental algorithm'' that calculates incremental
differences, as described in \citet{west1979}.

@@

The primary purpose of the noise component, \model{NOI} is to measure
the noise in each bolometer residual once all of the other models have
been fit and removed. This noise may then be used to estimate weights
for the data in subsequent iterations.

As in Section~\ref{sec:psd} the FFT of each bolometer time series is
taken, and the PSD calculated. An average white noise level is then
measured from 2 to 10\,Hz, a relatively clean region that lies above
the $1/f$ knee, but below the high-frequency line features for most of
the bolometer data taken during S2SRO (Fig.~\ref{fig:pspec}). Taking
this constant level for $\mathbf{P}(f)$ we then calculate the expected
variance of the time-series (in approximately 200\,Hz samples) using
Eq.~\ref{eq:psd}. If the bolometer noise were produced purely by
uncorrelated thermal sources (i.e. no other long-scale drifts), with
no high-frequency line features, and without the attenuation at even
higher frequencies by the anti-aliasing filter, this is the
theoretical noise limit of the detectors. These measured variances are
stored, and then used in subsequent iterations to weight the data
points when calculating the map estimate (Section~\ref{sec:ast}).

Since noise estimation is usually calculated as the final step, it is
the cleanest data for which some cleaning operations may be
performed. For this reason \model{NOI} can optionally use the
time-domain DC step fixer (Section~\ref{sec:steps}) and spike
detection (Section~\ref{sec:timedespike}).

@@

\subsection{Model degeneracies}

A fundamental problem faced by the SMURF algorithm, as described, is
that there are degeneracies between model parameters. Thinking of each
pixel in the output map, as well as each sample in the various noise
model components, such as the common-mode and signal removed by the
Fourier filtering, any degeneracy can cause the solution to
diverge. Of particular concern are situations in which the map
diverges.

>>>>

% journals
\newcommand{\ao}{\rm AO}
\newcommand{\apj}{\rm ApJ}
\newcommand{\apjl}{\rm ApJL}
\newcommand{\apjs}{\rm ApJS}
\newcommand{\aaps}{\rm A$\&$AS}
\newcommand{\aap}{\rm A$\&$A}
\newcommand{\aapr}{\rm A$\&$AR}
\newcommand{\mnras}{\rm MNRAS}
\newcommand{\aj}{\rm Astron. J.}
\newcommand{\araa}{\rm ARAA}
\newcommand{\nat}{\rm Nature}
\newcommand{\pasj}{\rm PASJ}
\newcommand{\pasp}{\rm Publ. Astron. Soc. Pac.}
\newcommand{\prd}{\rm PhRvD}
\newcommand{\ASP}{\rm ASP Conference Series}
\newcommand{\CASP}{\rm Comm. Astrophys. Space Phys.}
\newcommand{\astroph}{\rm astro-ph/}

@@

\author[Edward~L.~Chapin~et~al.]{
  \parbox[t]{\textwidth}{
    Edward~L.~Chapin$^{1,2,3}$\thanks{E-mail:~echapin@sciops.esa.int},
    David~S.~Berry$^{3}$,
    Andrew~G.~Gibb$^{2}$,
    Tim~Jenness$^{3}$,
    Douglas~Scott$^{2}$,
    Frossie~Economou$^3$\thanks{Present address: National Optical
      Astronomy Observatory, 950 N.\ Cherry Avenue, Tucson, AZ 85719, USA}
  }
  \\
  \\
  $^{1}$XMM SOC, ESAC, Apartado 78, 28691 Villanueva de la Ca\~nada, Madrid,
  Spain\\
  $^{2}$Dept. of Physics \& Astronomy, University of British Columbia,
  6224 Agricultural Road, Vancouver, B.C. V6T 1Z1, Canada\\
  $^{3}$JointAstronomy Centre, 660 N. A`oh\={o}k\={u} Place, University
  Park, Hilo, Hawaii 96720, USA}

@@

%\textbf{Guideline for figures:}

%\begin{itemize}
%\item minimize excess whitespace around the figure to optimize usage of space
%\item ensure a decent line thickness
%\item use a Times font for labels with a thickness that resembles the
%  text in the body. The font size should be similar to the main body
%  text or larger
%\item avoid using colour unless it is necessary for emphasis
%\end{itemize}

%\textbf{start intro here...}

@@

Over the last twenty years, observations in the submillimetre band
(defined here to be 200--1000\,\micron) have helped revolutionize
several important areas of astrophysics, including: the
characterization of early stages of star-formation by identifying the
dense, cold regions in molecular clouds where stars may eventually
form; discovering through blind surveys a class of massive dusty
star-forming galaxies in the early ($z>2$) Universe, now referred to
as submillimetre galaxies, or SMGs; and identifying debris disks
around nearby stars, the early stages of planet formation.  With
$\sim10000$ total detectors, SCUBA-2 is presently the largest
submillimetre camera in the world, and the fastest wide-area
ground-based submillimetre imager.

Most submillimetre imaging cameras, to date, have used bolometers to
maximize sensitivities \citep[e.g.][]{griffin2002}. Each pixel
consists of a thermal absorber to collect photons that have passed
through broad-band filters, and the signal is read-out using a
thermometer. The high-frequency broad-band noise in modern systems,
such as SCUBA-2, is limited by photon noise from the instrument and
ambient backgrounds. The low-frequency noise, however, is dominated by
sources such as slow variations in mean backgrounds (such as thermal
variations within the cryostat, and the atmosphere for ground-based
cameras), and drift in the readout electronics. Since the
low-frequency noise is largely correlated between all, or subsets, of
the bolometers in time, it can be suppressed during map-making, since
astronomical signals have the distinct property that they are fixed in
a sky reference frame (assuming they are not time-varying). If
successful, the noise in the resulting map is said to be ``background
noise limited'', meaning that the only significant remaining noise
sources are uncorrelated spatially in the map (the theoretical limit).

There are numerous ways to attack this problem, both in terms of the
data-collection method, and data processing. Historically, both radio
and submillimetre telescope systems have used ``chopping'' to perform
rapid differences between target sky locations and some reference
(e.g., another point on the sky, internal calibrators, and different
frequencies in the case of heterodyne systems). In fact, spatial
chopping was used by SCUBA, the predecessor to SCUBA-2, to remove
low-frequency drifts \citep{holland1999}. However, chopping (in
particular spatial chopping) has the unfortunate side-effect of
filtering out information on scales larger than the chop-throw
\textbf{(references)}.For this reason, modern cameras like SCUBA-2 aim
to provide a high-degree of inherent detector stability, and rapidly
slew the telescope so that astronomical signals up to large scales are
captured in a signal band at considerably higher frequency than the
residual low-frequency drift.

In the case of chopped data, ideally the resulting data are completely
free of low-frequency noise, and therefore the data samples are
statistically uncorrelated.

often used various
``chopping'' methods to remove low-frequency drifts -- rapid
differences are made between the sky signal at a given position and an
internal reference signal; the sky signal in two different bands for
heterodyne systems (frequency chopping); and a different position in
the sky \citep[a method employed by SCUBA, the predecessor of
SCUBA-2][]{holland1999}.

 many of them evolving
out of techniques for producing maps of the large-scale anisitropies
in the Cosmic Microwave Background (CMB)

  Styles of map-making: least
squares \citep{janssen1992} maximum likelihood
\citep[e.g.,][]{patanchon2008}; iterative -- see \citet{johnstone2000}
implementation of \citet{wright1996} to remove chop,
\citet{kovacs2008} for SHARC-2 etc.; de-correlation using PCA analysis
\citep[e.g.][]{laurent2005,scott2008,aguirre2010}.

@@

We then express the noise, $\mathbf{n}_i(t)$, as the sum of several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) + \mathbf{n}^\mathrm{f}_i(t) +
  \mathbf{n}^\mathrm{c}(t),
\label{eq:noise}
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white (thermal)
noise, $\mathbf{n}^\mathrm{f}_i(t)$ is low-frequency noise that is
\emph{not} correlated from bolometer-to-bolometer, and
$\mathbf{n}^\mathrm{c}(t)$ is a (predominantly low-frequency)
correlated or \emph{common-mode} component. The primary goal of
map-making is to model and remove $\mathbf{n}^\mathrm{f}_i$ and
$\mathbf{n}^\mathrm{c}$ from the time-series, while re-gridding the
remaining data we can then hope to approach the theoretical noise
limit in the map. This procedure is summarized in Fig.~\ref{fig:dimm}.

@@

In addition, a variance map $\mathbf{v}(x_j,y_j)$ is estimated. The
default behaviour is to estimate this weighted error on the mean given
the scatter in the weighted samples. This is accomplished by dividing
the \emph{biased weighted sample variance} by the number of samples
that went into the average (akin to the formula for standard error on
the mean, but accounting for weights),
%
\begin{equation}
\label{eq:varmap}
\mathbf{v}(x_j,y_j) = \frac{\sum_j \mathbf{w}_j
                            \sum_j \mathbf{w}_j \mathbf{b}_j -
                            \left( \sum_j \mathbf{w}_j \mathbf{b}_j \right)^2 }
                           { N \left( \sum_j \mathbf{w}_j \right)^2 },
\end{equation}
%
where $N_j$ is the total number of bolometer samples that land in the
pixel. As written, this algorithm is numerically unstable if the two
terms in the numerator are large, and of nearly the same value:
floating point precision errors can cause the difference to be
significantly incorrect. In practice we use the more numerically
stable ``weighted incremental algorithm'' that calculates incremental
differences, as described in \citet{west1979}.

@@

The primary purpose of the noise component, \model{NOI}, is to measure
the white noise of each bolometer ($\mathbf{n}^\mathrm{w}_i(t)$ in
Eq.~\ref{eq:noise}). The measurement generally occurs once all of the
other models have been fit and removed. This noise may then be used to
estimate weights for the data in subsequent iterations.

As in Section~\ref{sec:psd} the FFT of each bolometer time series is
taken, and the PSD calculated. An average white noise level is then
measured from 2 to 10\,Hz, a relatively clean region of the PSD that
tends to lie above the $1/f$ knee, but below the high-frequency line
features for typical bolometer data (Fig.~\ref{fig:pspec}). Taking
this constant level for $\mathbf{P}(f)$ we then calculate the expected
variance of the time-series (in approximately 200\,Hz samples) using
Eq.~\ref{eq:psd}. If the bolometer noise were produced purely by
uncorrelated thermal sources (i.e. no other long-scale drifts), with
no high-frequency line features, and without the attenuation at even
higher frequencies by the anti-aliasing filter, this is the
theoretical noise limit of the detectors. These measured variances are
stored, and then used in subsequent iterations to weight the data
points when calculating the map estimate (Section~\ref{sec:ast}). They
are also used for convergence tests (Section~\ref{sec:converge}).

Since noise estimation is usually calculated as the final step in the
iteration, it is the cleanest data for which some cleaning operations
may be performed. For this reason \model{NOI} can optionally use the
time-domain DC step fixer (Section~\ref{sec:steps}) and spike
detection (Section~\ref{sec:timedespike}).

Finally, it should be noted that the default behaviour is to calculate
the white noise levels \emph{once} within \emph{NOI}, after the second
iteration. The reason for fixing these values is to prevent any
potential divergence in the weight estimates with iterations, and also
to provide a fixed reference for the convergence tests
(Section~\ref{sec:converge}). Note that the absolute values of the
noise, and therefore weights calculated by the model component are
irrelevant (only their relative values). The reason is that the final
noise in the map is measured empirically from the scatter of the data
points that land in each pixel (Section~\ref{sec:ast}), and
large-scale noise sources can be checked by analyzing the angular
power spectra of jackknife maps (e.g., Section~\ref{sec:cosmo}).

@@

\subsection{Convergence tests and model degeneracies}
\label{sec:converge}

The map-maker will halt if the convergence criteria have been
met. Presently there are two numerical tests that may be performed by
SMURF, or alternatively, a fixed number of iterations may be
specified. The first numerical test is an approximate check of the
change in reduced chi-squared, $\chi^2_\mathrm{r}$. The standard
deviations, $\sigma_i$, are measured for each of the residual
bolometer signals once the modeled signal components have been
removed. In the early iterations, the residual signals usually contain
both white noise, and other long-timescale features. However, once the
solution has converged, this signal should look approximately
white. Once the white nosie, $\mathbf{n}^\mathrm{w}_i(t)$, has been
measured from the bolometer PSDs in \model{NOI}
(Section~\ref{sec:noi}), $\chi^2_\mathrm{r}$ is calculated as $(1/N)
\sum_i \sigma^2_i / \mathbf{n}^\mathrm{w}_i(t)$, where $i$ runs over
the $N$ bolometers. In other words, it is the average ratio between
the measured bolometer variances and their theoretical white-noise
levels. Note that this expression \emph{does not account for the
degrees of freedom}. Clearly there are a large number of parameters in
the model, which will ``fit-out'' some of the uncorrelated white
noise, and potentially bias this estimate of $\chi^2_r$ low. One
approach would be to keep track of the degrees of freedom, and
therefore provide a correction, as in \citet{kovacs2008}. However,
note that we are only using this value as a convergence test, and its
value is not used in an absolute sense. Furthermore, the reference
white noise values are calculated only once in SMURF
(Section~\ref{sec:noi}), and the number of model parameters are fixed,
so the value of $\chi^2_r$ should decrease monotonically if the
solution is behaving well. The convergence criterion is met if the
change in subsequent values of $\chi^2_r$ is smaller than the
parameter \texttt{chitol}, which is set to $10^{-3}$ by default.

It was immediately found in the analysis of SCUBA-2 data that the
$\chi^2_r$ test described above was not a sufficient stopping
criterion. The reason for this is that various components of the model
are, under normal circumstances, highly degenerate. Once simple
example is the degeneracy between large-scale astronomical structures
(larger than the array footprint), and the common-mode rejection step
(\model{COM}), which will be illustrated in
Section~\ref{sec:point}. Since the flatness of the residual only
reflects the ability of the model to fit the shape of the bolometer
time-series, it does not necessarily correlate with convergence of the
most important model component: the map. Indeed, examination of the
map after large numbers of iterations (e.g., $>10$), shows the
presence of large-scale structures that are anti-correlated with
structures in other model components. For this reason, a second
map-based convergence test was added.

The map-based convergence statistic, $M_c$, is the average absolute
change in the value of map pixels between subsequent iterations,
normalized by the map pixel uncertainties (square root of
Eq.~\ref{eq:varmap}), or
%
\begin{equation}
M^j_c = \frac{1}{N} \sum_i \frac{| \mathbf{m}(x_j,y_j) -
  \mathbf{m}(x_{j-1},y_{j-1}) |} {\sqrt{\mathbf{v}(x_j,y_j)}} ,
\end{equation}
%
where $i$ runs over the $N$ map pixels, and $j$ enumerates
iteration. The convergence criterion is met once this average change
is smaller than the parameter \texttt{maptol}. Experimentally we found
that a threshold of 0.05 gives good results (on average, map pixels
change by 5\% of the estimated map RMS in subsequent iterations).
This provides good correspondance with what we would choose ``by
eye'', and furthermore, letting the solver run for many more
iterations in several test cases yields insignificant differences.
